---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r set_globals, echo=FALSE, results="hide"}
# Do echo the code, and show the results.
knitr::opts_chunk$set(echo = TRUE, results = "show")
```

## Loading and preprocessing the data

```{r}
df <- read.csv(unz("activity.zip", "activity.csv"), stringsAsFactors = FALSE)
```

## What is mean total number of steps taken per day?

```{r}
# 1. Make a histogram of the total number of steps taken each day

# Aggregate steps per day (excluding NA values)
daily_steps <- aggregate(steps ~ date, data = df, FUN = sum, na.rm = TRUE)

hist(daily_steps$steps,
     main = "Histogram of Total Daily Steps",
     xlab = "Total Steps Taken per Day",
     col = "skyblue",
     border = "white")
```

```{r}
# 2. Calculate and report the **mean** and **median** total number of steps taken per day

mean_steps <- mean(daily_steps$steps, na.rm = TRUE)
cat("Mean steps: ", mean_steps)

median_steps <- median(daily_steps$steps, na.rm = TRUE)
cat("Median steps: ", median_steps)
```

## What is the average daily activity pattern?

```{r}
# 1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

mean_interval_steps <- aggregate(steps ~ interval, data = df, FUN = mean, na.rm = TRUE)

plot(mean_interval_steps$interval, mean_interval_steps$steps,
     type = "l",                             # line plot
     xlab = "5-minute Interval",
     ylab = "Average Steps",
     main = "Average Daily Activity Pattern",
     col = "skyblue")
```

```{r}
# 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

max_steps <- mean_interval_steps[which.max(mean_interval_steps$steps), ]
cat("Max average steps occur at interval", max_steps$interval,
    "with", round(max_steps$steps, 2), "steps on average.\n")
```

## Imputing missing values

```{r}
# 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s)

cat("Number of rows with missing steps", sum(is.na(df$steps)))
```

```{r}
# 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

# Impute the missing steps, using the average number of steps for that 5-minute interval across all days.
# - This approach assumes that the activity pattern for a given interval is relatively consistent day to day.

avg_interval_steps <- aggregate(steps ~ interval, data = df, FUN = mean, na.rm = TRUE)
```


```{r}
# 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

# Merge average steps per interval back to the original data
df_filled <- merge(df, avg_interval_steps, by = "interval", suffixes = c("", "_avg"))

# Replace NA steps with the average for that interval
df_filled$steps[is.na(df_filled$steps)] <- df_filled$steps_avg[is.na(df_filled$steps)]

# Drop the extra average column
df_filled$steps_avg <- NULL
```

```{r}
# 4. Make a histogram of the total number of steps taken each day.

daily_steps_filled <- aggregate(steps ~ date, data = df_filled, FUN = sum, na.rm = TRUE)

hist(daily_steps_filled$steps,
     main = "Histogram of Total Steps per Day (missing steps have been filled)",
     xlab = "Total Steps",
     col = "skyblue",
     border = "white")

```{r}
# Calculate and report the **mean** and **median** total number of steps taken per day.

mean_steps_filled <- mean(daily_steps_filled$steps, na.rm = TRUE)
cat("Mean steps (filled): ", mean_steps_filled)

median_steps_filled <- median(daily_steps_filled$steps, na.rm = TRUE)
cat("Median steps (filled): ", median_steps_filled)

# - Do these values differ from the estimates from the first part of the assignment?
mean_diff <- mean_steps_filled - mean_steps
cat("Mean steps (filled vs missing): ", mean_steps_filled,  mean_steps, " difference:", mean_diff)

median_diff <- median_steps_filled - median_steps
cat("Median steps (filled vs missing): ", median_steps_filled,  median_steps, " difference:", median_diff)

# - What is the impact of imputing missing data on the estimates of the total daily number of steps?

diff_check <- function(diff, name)
{
  epsilon <- 0.001
  if (diff > 0)
  {
    cat("INCREASED: estimate of *", name, "* total daily number of steps.")
  } else if (abs(diff) < epsilon)
  {
    cat("ABOUT SAME: estimate of *", name, "* total daily number of steps.")
  } else {
    cat("DECREASED: estimate of *", name, "* total daily number of steps.")
  }
}
diff_check(mean_diff, "Mean")
diff_check(median_diff, "Median")
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# 1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

df_filled$date <- as.Date(df_filled$date)

# Get the day of the week
day_type <- weekdays(df_filled$date)

# Classify as "weekday" or "weekend"
df_filled$day_type <- ifelse(day_type %in% c("Saturday", "Sunday"), "weekend", "weekday")

# Convert to a factor
df_filled$day_type <- factor(df_filled$day_type, levels = c("weekday", "weekend"))
```

```{r}
# Compute overall Y-axis range for both plots
y_range <- range(df_filled$steps, na.rm = TRUE)

# Set up 2-panel plot
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

# Weekday plot
with(subset(df_filled, day_type == "weekday"), {
  plot(interval, steps, type = "l", col = "blue",
       main = "Weekday: Average Steps by 5-Minute Interval",
       xlab = "Interval", ylab = "Average Steps",
       ylim = y_range)  # Set consistent Y-axis
})

# Weekend plot
with(subset(df_filled, day_type == "weekend"), {
  plot(interval, steps, type = "l", col = "darkgreen",
       main = "Weekend: Average Steps by 5-Minute Interval",
       xlab = "Interval", ylab = "Average Steps",
       ylim = y_range)  # Set consistent Y-axis
})
```

### Observations from the plots:
- weekdays show a broader active time range (from earlier in the day, to later in the day).
- weekdays generally show a higher average number of steps, especially from 5-minute intervals ~550 to ~1500
  - exception: *weekends* show a higher average number of steps from 5-minute intervals ~1600 to ~1700
- note: here, ~ indicates 'approximately'
